---
title: "Data Wrangling with the tidyverse – Post-Workshop Notes"
author: "Cecilia Baldoni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    code_folding: show
---

# Introduction

This document accompanies the online workshop **"Data Wrangling with the tidyverse"**, delivered as part of the summer school series for R beginners.

In this session, we explored how to:

- Inspect and explore data
- Select and filter rows/columns
- Create new variables and modify existing ones
- Summarize data with grouping
- Create basic plots using `ggplot2`

This R Markdown file contains all code used during the session, along with brief explanations and suggested follow-up questions. You can re-run the code, modify it, or expand upon it to reinforce your learning.

---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gapminder)
```

---

# Exercise 1: Loading & Exploring Data

Let's start by looking at the dataset.

```{r}
head(gapminder)
glimpse(gapminder)
summary(gapminder)
gapminder %>% distinct(continent)
gapminder %>% distinct(year)
gapminder %>% count(continent)
```

**Questions:**
- How many variables are there?
- What are the variable types?
- What years are included in the dataset?

---

# Exercise 2: Selecting and Filtering Data

We'll choose a subset of the columns and filter for recent European records.

```{r}
gapminder_selected <- gapminder %>%
  select(country, continent, year, lifeExp, gdpPercap)

gapminder_selected %>%
  filter(continent == "Europe", year > 2000) %>%
  arrange(desc(lifeExp))
```

**Questions:**
- Why might we want to focus on specific columns?
- Which countries have the highest life expectancy?

---

# Exercise 3: Creating & Modifying Variables

We now add a new variable for total GDP and categorize life expectancy.

```{r}
gapminder_gdp <- gapminder %>%
  mutate(gdp_total = gdpPercap * pop)

gapminder_renamed <- gapminder_gdp %>%
  rename(GDP_per_capita = gdpPercap)

gapminder_categorized <- gapminder_renamed %>%
  mutate(life_category = case_when(
    lifeExp < 50 ~ "Low",
    lifeExp < 70 ~ "Medium",
    TRUE ~ "High"
  ))

head(gapminder_categorized)
```

**Questions:**
- How are the categories defined?
- What other useful derived variables could we add?

---

# Exercise 4: Summarizing & Grouping Data

Let’s summarize life expectancy by continent and count countries.

```{r}
gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarise(mean_lifeExp = mean(lifeExp))

gapminder %>%
  group_by(continent) %>%
  summarise(country_n = n_distinct(country))
```

**Questions:**
- Which continent had the highest life expectancy in 2007?
- Why is grouping important for comparing data?

---

# Bonus: Visualizing Data

We use `ggplot2` to make three quick visual summaries.

```{r}
gapminder_2007 <- gapminder %>%
  filter(year == 2007)

# Scatter plot
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_x_log10() +
  labs(title = "GDP per Capita vs Life Expectancy (2007)",
       x = "GDP per Capita (log scale)", y = "Life Expectancy")

# Boxplot
ggplot(gapminder_2007, aes(x = continent, y = lifeExp, fill = continent)) +
  geom_boxplot() +
  labs(title = "Life Expectancy by Continent (2007)",
       x = "Continent", y = "Life Expectancy")
# Line plot
gapminder %>%
  group_by(year, continent) %>%
  summarise(mean_lifeExp = mean(lifeExp)) %>%
  ggplot(aes(x = year, y = mean_lifeExp, color = continent)) +
  geom_line(size = 1.2) +
  labs(title = "Average Life Expectancy Over Time by Continent",
       x = "Year", y = "Mean Life Expectancy")
```

---

# Further Learning

- [R for Data Science](https://r4ds.had.co.nz/)
- [Tidyverse Cheatsheet (PDF)](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf)
- [Gapminder Data](https://www.gapminder.org/data/)

---

# Session Info

```{r}
sessionInfo()
```

---

# Optional Challenge

Try applying what you learned to:
- Create a new variable for population in millions.
- Summarize GDP by year for Asia.
- Filter countries with life expectancy under 60 in 1997.

Happy wrangling!
